\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-11 doc/publications/plos16/plos16.tex
changes to plos16.tex

\newcommand{\cursedforest}{\textsc{CursedForest}\ }
\newcommand{\mtry}{{\texttt mtry\ }}
\newcommand{\ntree}{{\texttt ntree\ }}

@@

%%local edits (rob)
\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
{\raggedright\footnotesize #1}}
\reversemarginpar
\setlength{\marginparwidth}{5cm}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{placeins}
\begin{document}
\vspace*{0.2in}

@@

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest\ is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest\ outperforms existing tools even on small datasets where multi-threading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest\ can build a large number of
trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.
\marginpar{check this. It also allows us to set mtry to a large number, important when p is large}

@@

In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent
$B,B^\prime$.
\marginpar{why is this true}  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.

@@

The current implementation of \cursedforest\ uses an ``information gained'' criteria for splitting. Let $f_q$ be the
\marginpar{this is wrong -- it used Gini}
fraction of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The entropy of the node is
$H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$ The expected information gain is the change in entropy $H$ from a prior state
$T$ to a state that takes some information $a$ as given:
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children). 
Variable importance is the sum of IG over all notes and all trees for that variable.

@@

\subsubsection{Synthetic data} 
\marginpar{this is wrong now}
Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
probabilities.

The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ then 
\begin{align}
y=& \prod_{i=1}^{i=5} {w_i x_i} + \epsilon \textrm{ for $x_i \neq 0$} 
\intertext{where}
\epsilon \sim & N(0, \sigma_\epsilon^2= 0.12)
\end{align}
The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree\ the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a very poor fit and still have an accurate ranking of variable
  importance;
\item \mtry\ -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
In this section we demonstrate two features of \cursedforest\, a variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
samples, as discussed in the methods section.

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
classification error and the rank-biased-overlap
\cite[]{Webber.et.al.2010} measure.

@@

Fig.~\ref{figure:synth1} illustrates the effect of varying  \mtry\ and \ntree\ on these two measures. 
Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows that depends strongly on \ntree\ 
whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is more dependent on \mtry.
\marginpar{this is not true anymore}

@@

\marginpar{this was done with n=100000. Needs to redone with n=2.5M}

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{ OOB (out of bag) classification error and Rank-biased overlap estimates for random
        forests with 100 trees trained on the synthetic dataset of 2.5M features and 5K samples.}}
    \label{figure:synth1}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
      \caption{OOB classification error} 
      \label{figure:out-of-bag-prediction-error-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/rbo.png}
      \caption{The  rank-biased-overlap measure.} 
      \label{figure:rbo-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft} 
      Each measure is shown
      as a function of \mtry\ and \ntree. Note that the plot colors have been chose so that the lighter color indicates
      the better result, that is, a lower OOB and a higher RBO.
    \end{flushleft}
  \end{adjustwidth}
\end{figure}

@@

@@

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
\textsc{Ranger}, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
of random forests by using a range of \mtry\ values.

@@

First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. Regardless of the
number of cores used, \cursedforest displays approximately linear dependency between the execution time and the \mtry
parameter (presented here as a fraction of the number of variables, Fig. \ref{figure:synthetictiming.a}

\cursedforest\ scales almost linearly with the number of CPU cores for medium values of \mtry\ fraction but for both
lower and higher values the performance degrades slightly (Fig \ref{figure:synthetictiming.b}). In the latter case the likely cause is
communication overhead (with lower \mtry\ values the proportion of time for parallelizable computation to the time for
internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
computational capacity.

Next we investigate \cursedforest\ scalability with regards to the size of data, by varying the number of variables and
sample for a fixed \mtry\ fraction of 0.25 and execution of 128 CPU cores. The raw results are presented in the table
[REF] and visualized in figure [ REF: Fig 2] below (please note log scale on the axes and the values on y axes are
expresses as trees per hour).

Generally, the number of trees per hour decreases with increased number of variables and samples sizes. Some
irregularities in the graph can be attributed to computation vs communication tradeoff. It is also worth noting that
keeping \mtry\ fraction constants results with higher \mtry\ values with growing number of variables, and this is what
drives the performance down rather that the increase of dataset size itself.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Scalablity of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.}}
    \label{figure:synthetictiming}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} 
      \caption{Time in seconds to build 100 trees for different \mtry\ fractions. } 
      \label{figure:synthetictiming.a} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png}
      \caption{Number of trees build per hour when using a growing number of CPU cores.}
    \label{figure:synthetictiming.b}
    \vspace{4ex}
  \end{subfigure} 
\end{adjustwidth}
\end{figure}

@@

\section{Conclusion}
We have demonstrated that using a different paralleization model can extend random forests to the case of an
extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
the variables are uninformative and have demonstrated the utility of the model for large GWAs data sets. By comparing
this implementation to other implementations (including those optimized for large data sets) we have demonstrated the
utility of this approach.

@@

>>>>

\newcommand{\cursedforest}{\textsc{CursedForest}\xspace}
\newcommand{\mtry}{\texttt{mtry}\xspace}
\newcommand{\ntree}{\texttt{ntree}\xspace}

@@

%%local edits (rob)
\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
{\raggedright\footnotesize #1}}
\reversemarginpar
\setlength{\marginparwidth}{5cm}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{placeins}
\usepackage{xspace} % see http://tex.stackexchange.com/questions/31091/space-after-latex-commands

\begin{document}
\vspace*{0.2in}

@@

\marginpar{make consistent capitalization for ranger, randomForest}

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) data sets. A Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. Using this facility, \cursedforest is capable of
parallelization the split for each node in a tree thereby handling millions of features. It can, for example, fit a
model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
variants~\cite{1KG2012}.  By also utilizing Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest\ outperforms existing tools even on small datasets where multi-threading generally performs
well. Harnessing the virtually unlimited capability to parallelize tasks, \cursedforest can build a large number of
trees and use a high \mtry value which leads to greater accurace (see section \ref{section:synthetic}). 
\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
We do not see that}

@@

In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent
$B,B^\prime$.
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.

@@

The current implementation of \cursedforest\ uses an ``Gini impurity''
criteria for splitting. Let $f_q$ be the fraction of items labelled
with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
\[
I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 − f_q ), 
\]
which is a minimum when all observatkions at the node are in the same class.

@@

\subsubsection{Synthetic data} 
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomics variation) randomly generated from uniform distribution with equal
probabilities.

The model paramters are $w_i = 1/\sqrt{2^{i-1}}$ for $i =1,\ldots 5$ then 
\begin{align*}
z=& \Sigma_{i=1}^{i=5} {w_i x_i} \\
\intertext{and}
\sigma_\epsilon^2 =& Var(z)(1- 0.125)/0.125 \\
\intertext{so}
y =&  z+\epsilon \textrm{ where } \epsilon \sim  N(0, \sigma_\epsilon^2).
\end{align*}
The dichotomous response is generated by thresholding $y$ at the $0.8$ quantile.

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate two features of \cursedforest\, a variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000
samples, as discussed in the methods section.

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB 
classification error and the rank-biased-overlap
\marginpar{what is wrong with the formatting of this citation?}
\cite[]{Webber.et.al.2010} measure.

@@

\marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
Fig.~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
the default. Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
values of \mtry and \ntree considered  whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the rbo measure is less
sensitive once we over a certain threshold.

@@

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{ out of bag classification error and Rank-biased overlap estimates}} 
    \label{figure:synth1}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/out-of-bag-prediction-error.png}
      \caption{OOB classification error} 
      \label{figure:out-of-bag-prediction-error-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/rbo.png}
      \caption{The  rank-biased-overlap measure.} 
      \label{figure:rbo-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft} 
      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
      the better result, that is, a lower OOB or a higher RBO.
    \end{flushleft}
  \end{adjustwidth}
\end{figure}

@@

\marginpar{this section is not one sentence? has it been lost?}

@@

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
\textsc{Ranger}, allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity
of random forests by using a range of \mtry values.

@@

First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. Regardless of the
number of cores used, \cursedforest displays approximately linear dependency between the execution time and \mtry
(Fig. \ref{figure:synthetictiming.a}).

\cursedforest scales almost linearly with the number of CPU cores for medium values of \mtry fraction but for both
lower and higher values the performance degrades slightly (Fig \ref{figure:synthetictiming.b}). In the latter case the likely cause is
communication overhead (with lower \mtry values the proportion of time for parallelizable computation to the time for
internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
computational capacity.

Next we investigate \cursedforest scalability with regards to the size of data, by varying the number of variables and
sample for a fixed \mtry fraction of 0.25 and execution of 128 CPU cores. The raw results are presented in the table
[REF] 
\marginpar{have we lost this table?}
and visualized in figure \ref{figure:synth2} below (please note log scale on the axes and the values on y axes are
expresses as trees per hour).

Generally, the number of trees per hour decreases with increased number of variables and samples sizes. Some
irregularities in the graph can be attributed to computation vs communication tradeoff. It is also worth noting that
keeping the \mtry fraction constant results in higher \mtry values with the growing number of variables, and this is what
drives the performance down rather that the increase of dataset size itself.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Scalability of the wide random forest on the synthetic dataset of 2.5M features and 5K samples.}}
    \label{figure:synthetictiming}
    \begin{subfigure}[b]{0.45\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/mtry_cpu.png} 
      \caption{Time in seconds to build 100 trees for different \mtry\ fractions. } 
      \label{figure:synthetictiming.a} 
      \vspace{4ex}
    \end{subfigure} 
\hfill
    \begin{subfigure}[b]{0.45\linewidth}
      \centering
      \includegraphics[totalheight=6cm]{./figs/cpu_mtry_trees_per_hour.png}
      \caption{Number of trees build per hour when using a growing number of CPU cores.}
    \label{figure:synthetictiming.b}
    \vspace{4ex}
  \end{subfigure} 
\end{adjustwidth}
\end{figure}

@@

\section{Conclusion}
We have demonstrated that using a different parallelization model can extend random forests to the case of an
extremely large number of variables. We have treated the case of variable selection in a $p >> n$ model where most of
the variables are uninformative and have demonstrated the utility of the model for large GWAs data sets. By comparing
this implementation to other implementations (including those optimized for large data sets) we have demonstrated the
utility of this approach.

@@

We note several things here:
\begin{itemize}
\item 
\end{itemize}
